Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-11 article.tex
edit manuscript and include interference references in places

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	% Need more context/bridging
	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1pdm influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, this does not significantly change the results presented below.

In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.

@@

To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$.
The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C.
Initially, \emph{i.e.} at time $t_0=0$, this distribution is by construction peaked around $f_0$.
If a large fraction of the trajectories keep increasing after  this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen.
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time.
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.
For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$.
This result is exactly what is expected in a population evolving in the \emph{absence} of selection.
% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
% \begin{equation}
%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
%     \begin{cases}
%     1-e^{-2sfN} & sfN \gg 1 \\
%     f & |sN| \ll 1 \\
%     e^{-2s(1-f)N} & -sN \gg 1
%     \end{cases}
% \end{equation}
% The result for the neutral case is intuitive:
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$.
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation.
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it. \pierre{``limited but clear'' instead of ``small''?}
In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.

@@

Two ways of categorising mutations, however, suggest some power to predict fixation.
In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants.
Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation.
In panel \textbf{D}, we separated mutations that appear more than once or only once in the reconstructed tree (see methods), and found that the former fix more often.
Panels \textbf{C} and \textbf{D} show that it is possible to gain some information on the chance of fixation of a particular mutation, as was done in panel \textbf{B}.
However, the predictive power remains small, with the ``top'' curves in panels \textbf{C\&D} being very close to the diagonal.

@@

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the idea that seasonal influenza virus evolution is driven by rapid directed positive selection.
One possible explanation for the weakly predictable behaviour of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}.
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

>>>>

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: For A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a consistent tendency to rise in frequency (see below).

@@

To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$.
The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C.
Initially, \emph{i.e.}, at time $t_0=0$, this distribution is by construction peaked around $f_0$.
If a large fraction of the trajectories keep increasing after  this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen.
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time.
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.
For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$.
This result is exactly what is expected in a population evolving in the \emph{absence} of selection.
% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
% \begin{equation}
%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
%     \begin{cases}
%     1-e^{-2sfN} & sfN \gg 1 \\
%     f & |sN| \ll 1 \\
%     e^{-2s(1-f)N} & -sN \gg 1
%     \end{cases}
% \end{equation}
% The result for the neutral case is intuitive:
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$.
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation.
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
For A/H1N1pdm, we find that new mutations fix moderatly but consistently more often then their current frequency.
In both cases, frequency is a strong predictor of fixation.

@@

In supplementary figure \ref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.

In addition to mutational load, we find that parallel evolution is predictive of fixation.
In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants.
Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation.
In panel \textbf{D}, we separated mutations that appear more than once or only once in the reconstructed tree (see methods), and found that the former fix more often.
Panels \textbf{C} and \textbf{D} show that it is possible to gain some information on the chance of fixation of a particular mutation, as was done in panel \textbf{B}.
However, the predictive power remains small, with the ``top'' curves in panels \textbf{C\&D} being very close to the diagonal.

@@

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the idea that seasonal influenza virus evolution is driven by rapid directed positive selection.
One possible explanation for the weakly predictable behaviour of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
Indeed, \citet{illingworth_components_2012} have shown that background fitness and interference have substantial effects on frequyency trajectories.

Nevertheless, we expect that signatures of intrinsic fitness benefits remain.
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}.
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.
